---
title: "What are GAMs?"
author: "Eric Pedersen (with material heavily borrowed from David Miller)"
date: "1000&ndash;1230 CST (1600&ndash;1830 UTC) July 30th, 2020"
output:
  xaringan::moon_reader:
    css: ['default', 'https://stackpath.bootstrapcdn.com/bootstrap/4.3.1/css/bootstrap.min.css', 'slides.css']
    lib_dir: libs
    nature:
      titleSlideClass: ['inverse','middle','left',my-title-slide]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "macros.js"
      ratio: '16:9'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = usethis::proj_path())

library('here')
library('mgcv')
library('gratia')
library('gamair')
library('ggplot2')
library('purrr')
library('mvnfast')
library("tibble")
library('gganimate')
library('cowplot')
library('tidyr')
library("knitr")
library("viridis")
library('readr')
library('dplyr')
library('gganimate')

opts_chunk$set(cache=TRUE, echo=FALSE)
```


## Overview

- A very quick refresher on GLMs
- What is a GAM?
- How do GAMs work? (*Roughly*)
- What is smoothing?
- Fitting and plotting simple models

---
# A (very fast) refresher on GLMs

---

##  What is a Generalized Linear model (GLM)?

Models that look like:

$y_i \sim Some\ distribution(\mu_i, \sigma_i)$

$link(\mu_i) = Intercept + \beta_1\cdot x_{1i} + \beta_1\cdot x_{2i} + \ldots$

---

##  What is a Generalized Linear model (GLM)?

Models that look like:

$y_i \sim Some\ distribution(\mu_i, \sigma_i)$

$link(\mu_i) = Intercept + \beta_1\cdot x_{1i} + \beta_1\cdot x_{2i} + \ldots$

 The average value of the response, $\mu_i$, assumed to be a linear combination of the covariates, $x_{ji}$, with an offset
---

##  What is a Generalized Linear model (GLM)?

Models that look like:

$y_i \sim Some\ distribution(\mu_i, \sigma_i)$

$link(\mu_i) = Intercept + \beta_1\cdot x_{1i} + \beta_1\cdot x_{2i} + \ldots$


 The model is fit (not really...) by maximizing the log-likelihood:


$\text{maximize}  \sum_{i=1}^n logLik (Some\ distribution(y_i))$

$\text{ with respect to } Intercept, \ \beta_1,\ \beta_2, \ ...$

---
##  With normally distributed data (for continuous unbounded data):


$y_i = Normal(\mu_i , \sigma_i)$

$Identity(\mu_i) = Intercept + \beta_1\cdot x_{1i} + \beta_1\cdot x_{2i} + \ldots$

```{r gaussplot, fig.width=12, fig.height=6}
set.seed(2) ## simulate some data...
dat <- tibble(x = seq(0,1, length = 100),
              y = rnorm(100, 3*x+2, 0.5))

mod <- glm(y~x, data= dat, family=gaussian)

p <- ggplot(dat,aes(y=y,x=x)) +
  geom_point() +
  geom_smooth(method= glm,formula=y~x, method.args=list(family = "gaussian"))+
  labs(title= paste0("True: Identity(mu) = 2 + 3*x, sigma = 0.5\n",
                     "Estimated: Identity(mu) = ",
                     round(coef(mod)[[1]],1)," + ",
                     round(coef(mod)[[2]],1),"*x, sigma = ",
                     round(summary(mod)$dispersion,2)))+
  theme_minimal(base_size = 20)
  
print(p)
```

---

##  With Poisson-distributed data (for count data):

$y_i = Poisson(\mu_i)$

$\mu_i = Intercept + \beta_1\cdot x_{1i} + \beta_1\cdot x_{2i} + \ldots$

```{r poisplot, fig.width=12, fig.height=6}
set.seed(2) ## simulate some data...
dat <- tibble(x = seq(0,1, length = 100),
              y = rpois(100, exp(3*x+2)))

mod <- glm(y~x, data= dat, family= poisson(link ="log"))

p <- ggplot(dat,aes(y=y,x=x)) +
  geom_point() +
  geom_smooth(method= glm,formula=y~x, method.args=list(family = "poisson"))+
  labs(title= paste0("True: ln(mu) = 2 + 3*x\n",
                     "Estimated: ln(mu) = ",
                     round(coef(mod)[[1]],1)," + ",
                     round(coef(mod)[[2]],1),"*x"))+
  theme_minimal(base_size = 20)
  
print(p)
```

---

# Why bother with anything more complicated?

---

## Is this linear?


```{r islinear, fig.width=12, fig.height=6}
set.seed(2) ## simulate some data...
dat <- gamSim(1, n=400, dist="normal", scale=0.2, verbose=FALSE)
dat <- dat[,c("y", "x0", "x1", "x2", "x3")]
p <- ggplot(dat,aes(y=y,x=x1)) +
      geom_point() +
      theme_minimal(base_size = 20)
print(p)
```

---

## Is this linear? Maybe?


```{r eval=FALSE, echo=TRUE}
lm(y ~ x1, data=dat)
```


```{r maybe, fig.width=12, fig.height=6}
p <- ggplot(dat, aes(y=y, x=x1)) + 
  geom_point() +
  theme_minimal(base_size = 20)

print(p + geom_smooth(method="lm"))
```


---

# What is a GAM?


