---
title: "Extending GAMs"
author: "David L Miller"
date: "August 12, 2020"
output:
  xaringan::moon_reader:
    css: ['default', 'https://stackpath.bootstrapcdn.com/bootstrap/4.3.1/css/bootstrap.min.css', 'slides.css']
    lib_dir: libs
    nature:
      titleSlideClass: ['inverse','middle','left',my-title-slide]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "macros.js"
      ratio: '16:9'
---

```{r setup, include=FALSE, cache=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE, dev = 'svg', echo = TRUE, message = FALSE, warning = FALSE,
                      fig.height=6, fig.width = 1.777777*6)
library('here')
library('mgcv')
library('gratia')
library('gamair')
library('ggplot2')
library('purrr')
library('mvnfast')
library("tibble")
library('gganimate')
library('cowplot')
library('tidyr')
library("knitr")
library("viridis")
library('readr')
library('dplyr')
library('gganimate')
library('xaringan')
library('sf')
## plot defaults
theme_set(theme_minimal(base_size = 16, base_family = 'Fira Sans'))
## constants
anim_width <- 1000
anim_height <- anim_width / 1.77777777
anim_dev <- 'png'
anim_res <- 200
```

# So far:

- Simple models
- Normally distributed response

These are good models, but there's lots more to see!


---

# New things

- distribution of the data
  - `family=` argument
- adding dimensions to your smooths
  - space and time


---

# Distributions

- `family=` argument in `gam()`
- see `?family` for a list
- what you'd expect, as for `lm()` and `glm()`
- most useful cases:
  - `binomial` ( yes/no, $y \in \{0, 1\}$ )
  - `poisson` ( counts, $y \in \{0, 1, 2, \ldots\}$ )
  - `Gamma` ( positive, $y>0$ )


---

# Special count distributions

- Poisson is often not adequate for "real" counts
- Assuming $\mathbb{E}(y) = \text{Var}(y)$ is usually incorrect
 
Other options?
 
- `quasipoisson` (count-ish, $y \geq 0$)
  - awkward to check, no likelihood
- `tw`/`Tweedie` ( $y \geq 0$)
  - great for CPUE-type data
- `nb`/`negbin` ( $y \geq 0$)
  - can be used but strong assumption (see next)

---

# Tweedie distribution

.pull-left[
```{r tweedie, echo=FALSE, fig.height=5, fig.width=5}
par(ps=17)
library(RColorBrewer)
library(mgcv)

# taken from ?mgcv::ldTweedie
y <- seq(1e-10,10,length=1000)
p <- seq(1.2, 1.6, by=0.2)
phi <- 1
fy <- exp(ldTweedie(y,mu=2,p=p[1],phi=phi)[,1])

rr <- brewer.pal(8,"Dark2")
  
plot(y, fy, type="n", ylim=c(0,1), main="Tweedie density as q changes",
     xlab="Count", ylab="Density")
for (i in 1:length(p)) {
  fy <- exp(ldTweedie(y, mu=2, p=p[i], phi=phi)[, 1])
  lines(y, fy, col=rr[i], lwd=2.5)
}

legend(x="topright", legend=p, fill=rr, inset=0.05)

```
(NB there is a point mass at zero not plotted)
]
.pull-right[
-  $\text{Var}\left(\text{count}\right) = \phi\mathbb{E}(\text{count})^q$
- Poisson is $q=1$ 
- We estimate $q$ and $\phi$ for `tw`
- We set $q$ and estimate $\phi$ for `Tweedie`
]

---

# Negative binomial distribution

.pull-left[
```{r negbin, echo=FALSE, fig.height=5, fig.width=5}
y<-seq(1,12,by=1)
disps <- seq(0.01, 1, len=5)

fymat <- matrix(NA, length(y), length(disps))

i <- 1
for(disp in disps){
  fymat[,i] <- dnbinom(y, size=disp, mu=5)
  i <- i+1
}

plot(range(y), range(fymat), type="n", ylab="Density", xlab="Count", cex.lab=1.5,
     main="")

rr <- brewer.pal(8,"Dark2")

for(i in 1:ncol(fymat)){
  lines(y, fymat[,i], type="l", col=rr[i], lwd=2)
}

legend(x="topright", legend=disps, fill=rr, inset=0.05)

```
]
.pull-right[
- $\text{Var}\left(\text{count}\right) =$ $\mathbb{E}(\text{count}) + \kappa \mathbb{E}(\text{count})^2$
- Poisson is $\kappa=0$
- Estimate $\kappa$ for `nb`
- Set $\kappa$ for `negbin` 
]

---

# 游游멇릟 Example 游냐游멇릟

.pull-left[
- Let's look at species richness over time
- Number of species in each trawl (counts!)

```{r richness}
rich <- read.csv(here("data/trawl_nl.csv"))
```
]

.pull-right[
```{r richness-violin, fig.height=5, fig.width=5, echo=FALSE}
ggplot(rich) +
  geom_violin(aes(richness, factor(year))) +
  labs(x="Number of species", y="Year")
```

]

---

# 游游멇릟 Example 游냐游멇릟

.pull-left[
- Does richness vary over time?
- Let's try assuming the response is Poisson!

```{r pois-rich-fit}
b_rich <- gam(richness~s(year), family=poisson,
              method="REML", data=rich)
```
]

.pull-right[
```{r plot-year-rich, fig.width=5}
plot(b_rich, shade=TRUE)
```
]

---

# `summary` output

```{r rich-sum, echo=FALSE}
summary(b_rich)
```

---

# Other `family` stuff

- can specify a `link=` argument
  - (usually don't have to)
  - `?family` has options for each distribution
- for a fitted model `model$family`
  - details of what was used
  - `model$family$linkfun()` gives link function
  - `model$family$linkinv()` gives inverse link
- More on checking distributions tomorrow!

---

class: inverse middle center big-subsection

## Beyond one dimensional smooths:<br/> space, time and more

---

# 游붏游붏 Example 游붏游붏

.pull-left[
```{r data-load}
shrimp <- read.csv(here("data/trawl_nl.csv"))
shrimp2010 <- subset(shrimp, year==2010)
range(shrimp2010$shrimp)
head(shrimp2010$shrimp, 3)
```

Not normal, not integer!

]

.pull-right[

```{r, fig.height=5, fig.width=5, echo=FALSE}
hist(shrimp2010$shrimp, xlab="Biomass", main="")
```

]

---

# 游붏游딬

.pull-left[
- spatial variation!
- how do we model this??
- `s(long) + s(lat)` misses the interaction
]

.pull-right[
```{r biom-space-plot, fig.height=11, echo=FALSE}
coast <- read_sf(here("data/nl_coast.shp"))

ggplot(shrimp2010) +
  geom_point(aes(x=long, y=lat, size=shrimp)) +
  geom_sf(data=coast) +
  coord_sf()
```
]

---

# Adding dimensions

- Thin plate regression splines (default basis)
- `s(x, y)`
- Assumes `x` and `y` are measured in the same units
  - `x`, `y` projected coordinates 游녨
  - `x` temperature, `y` depth 游녩

---

# 游붏游딬

Trying out quasi-poisson response


.pull-left[
```{r shrimpmod}
b_shrimp <- gam(shrimp ~ s(x, y),
                data=shrimp2010,
                family=quasipoisson,
                method="REML")
```
]

.pull-right[
```{r plot-effect, fig.height=8}
plot(b_shrimp, asp=1)
```
]

---

# That plot is hard to understand!

.pull-left[
```{r plot-effect-bad, fig.height=10}
plot(b_shrimp, asp=1)
```
]

.pull-right[
```{r plot-effect-good, fig.height=10}
plot(b_shrimp, scheme=2, asp=1)
```
]

---

# `summary` output

```{r summary-spat, echo=FALSE}
summary(b_shrimp)
```

---

class: inverse middle center big-subsection

## Now give it a try!


---

class: inverse middle center big-subsection

## Other multi-dimensional smooths

---

# `s(x,y)` doesn't always work

- Only works for `bs="tp"` or `bs="ts"`
- Covariates are isotropic
- What if we wanted to use lat/long?
- Or, more generally: interactions between covariates?

---

# Enter `te()`

.pull-left[
- We can built interactions using `te()`
- Construct 2D basis from 2 1D bases
- Biomass as a function of temperature and depth?
  - `te(temp_bottom, depth)`
- 游눬 "marginal 1Ds, join them up"
]

.pull-right[
```{r tensor, echo=FALSE, results='hide', fig.width=8, fig.height=6, messages=FALSE, warning=FALSE}
b_shrimp_ll_ti <- gam(shrimp ~ ti(temp_bottom, depth) + ti(temp_bottom) + ti(depth),
                data=shrimp2010,
                family=quasipoisson,
                method="REML")

layout(matrix(c(1,2,3,3), 2, 2), widths=c(1.5,1.5,1), height=c(1,1,2))
opar <- par(mar=c(4, 3, 1, 2) + 0.1)
plot(b_shrimp_ll_ti, select=3)
plot(b_shrimp_ll_ti, select=2)
par(mar=c(0, 0, 0, 0) + 0.1)
vis.gam(b_shrimp_ll_ti, view=c("temp_bottom","depth"), theta=-60, phi=30, color="bw")
par(opar)
```
]

---

# Using `te()`

Just like `s()`:


```{r te-ex}
shrimp_te <- gam(shrimp ~ te(temp_bottom, depth),
                data=shrimp2010,
                family=quasipoisson,
                method="REML")
```

---

# `summary`

```{r te-ex-summary, echo=FALSE}
summary(shrimp_te)
```

---

# Things to fiddle with

- Setting `k=` 2 ways:
  - `k=5`: 5 for all covariates (total $5*5=25$)
  - `k=c(3,5)`: per basis, in order (total $3*5=15$)
- Setting `bs=` 2 ways:
  - `bs="tp"`: tprs for all bases
  - `bs=c("tp", "tp")`: tprs per basis

---

# Pulling `te()` apart: `ti()`

- Can we look at the components of the `te()`
- `te(x, y) = ti(x, y) + ti(x) + ti(y)`

```{r tiex}
shrimp_ti <- gam(shrimp ~ ti(temp_bottom, depth) +
                          ti(temp_bottom) + ti(depth),
                data=shrimp2010,
                family=quasipoisson,
                method="REML")
```

---

# `summary`

```{r ti-ex-summary, echo=FALSE}
summary(shrimp_ti)
```

---

class: inverse middle center big-subsection

## Now give it a try!


---
class: inverse middle center big-subsection

## Building spatio-temporal models

---

# 游붏游딬游딉

```{r biom-space-time-plot, fig.height=8, fig.width=15, echo=FALSE}
ggplot(shrimp) +
  geom_point(aes(x=long, y=lat, size=shrimp), alpha=.5) +
  geom_sf(data=coast) +
  facet_wrap(~year, ncol=5)
```


---

# Space x time

- We had a 2d spatial model, add time?
  - `te(x,y,year)` ?
- Want that 2d smooth rather than `te(lat, long)`?
- `d=` groups covariates
  - `te(x, y, year, d=c(2,1))` gives `x,y` smooth and `year` smooth tensor
- Assuming default `k=` and `bs=` for bases

---

# Fiddling

- Often fewer temporal replicates
  - Fewer years than unique locations
  - `k=` smaller for temporal covariate?
- Use cubic spline basis for time?
  - simpler basis, even knot placement
- When using `ti()` everything needs to match up!

---

# 游붏游딬游딉

Putting that together:

```{r ti-xyt}
shrimp_xyt <- gam(shrimp ~ ti(x, y, year, d=c(2,1), 
                              bs=c("tp", "cr"), k=c(20, 5)) +
                           ti(x, y, d=2, bs="tp", k=20) +
                           ti(year, d=1, bs="cr", k=5),
                data=shrimp,
                family=quasipoisson,
                method="REML")
```

---

# `summary`

```{r xyt-summary, echo=FALSE}
summary(shrimp_xyt)
```

---

# Total effects

```{r ti-xyt-plot, echo=FALSE, results="hide"}
par(mfrow=c(3,4))
lapply(2005:2014, function (t) vis.gam(shrimp_xyt, view=c("x", "y"), cond=list(year=t),
       plot.type="contour", asp=1, main=t, too.far=0.1))
```

---

# `ti(x, y, year)`

```{r ti-xyt-plot-xyt, echo=FALSE, results="hide", fig.width=7}
plot(shrimp_xyt, select=1, scheme=2, asp=1)
```

---

# `ti(x, y)` and `ti(year)`

```{r ti-xyt-plot-xy-t, echo=FALSE, results="hide", fig.width=7}
par(mfrow=c(1, 2))
plot(shrimp_xyt, select=2, scheme=2, asp=1)
plot(shrimp_xyt, select=3, scheme=2, scale=0, seWithMean=TRUE)
```

---
class: inverse middle center big-subsection

## One more thing:<br/>centring constraints

---

# Adding terms to a model

.pull-left[
- Already know `s(x)` has "all the bits"
- basis functions, intercept:
- with multiple terms, can't have an intercept for each
- identifiability
]

.pull-right[
```{r centring-ex, echo=FALSE, results="hide", fig.width=7}
plot(shrimp_xyt, select=3, scheme=2, scale=0)
```



]
---
  
# What does the intercept mean?

.pull-left[
- Solution: absorb all the intercepts into one: $\beta_0$
- 游눬 "base level of a factors"
- shared over all smooths

```{r}
coef(shrimp_xyt)[1]
```
]

.pull-right[
```{r centring-ex3, echo=FALSE, results="hide", fig.width=7}
plot(shrimp_xyt, select=3, scheme=2, scale=0)
```
]

---

# Thinking further

- Need to avoid identifiability issues: use constraints
- Ensure that mean value $s(x)$ is 0
- (Shape isn't changed, just a vertical, *intercept*, shift)
- Sometimes leads to surprises in your results, compared to their specification

---
class: inverse middle center big-subsection

## Recap

---

# What did we learn?

- set response distribution
  - `family=` argument
  - see `?family`
- spatial smoothing
  - `s(x,y)` for projected coordinates
  - `te(lat, long)` for latitude and longitude (better to project?)
- interactions
  - `te(covar1, covar2, ...)`
  - `ti()` to decompose the effects
- space and time
  - `te(x, y, time, d=c(2,1))`
  - again `ti()` to decompose

