---
title: "Prediction & Model Uncertainty"
author: "Gavin L. Simpson"
date: "August 13, 2020"
output:
  xaringan::moon_reader:
    css: ['default', 'https://stackpath.bootstrapcdn.com/bootstrap/4.3.1/css/bootstrap.min.css', 'slides.css']
    lib_dir: libs
    nature:
      titleSlideClass: ['inverse','middle','left',my-title-slide]
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "macros.js"
      ratio: '16:9'
---

```{r setup, include=FALSE, cache=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(cache = TRUE, dev = 'svg', echo = TRUE, message = FALSE, warning = FALSE,
                      fig.height=6, fig.width = 1.777777*6)
library('here')
library('mgcv')
library('gratia')
## library('gamair')
library('ggplot2')
library('sf')
## library('purrr')
## library('mvnfast')
## library("tibble")
## library('cowplot')
## library('tidyr')
## library("knitr")
## library("viridis")
## library('readr')
library('dplyr')
## library('gganimate')

## plot defaults
theme_set(theme_minimal(base_size = 16, base_family = 'Fira Sans'))
## constants
anim_width <- 1000
anim_height <- anim_width / 1.77777777
anim_dev <- 'png'
anim_res <- 200
```

# Recap &mdash; shrimp species richness

.row[
.col-6[
```{r load-shrimp}
shrimp <- read.csv(here('data', 'trawl_nl.csv'))
```

```{r shrimp-richness}
m_rich <- gam(richness ~ s(year),
              family = poisson,
              method = "REML",
              data = shrimp)
```
]
.col-6[
```{r richness-violin, fig.height=5, fig.width=5, echo=FALSE}
ggplot(shrimp) +
  geom_violin(aes(x = richness, y = factor(year))) +
    labs(x = "Number of species", y = "Year")
```
]
]

---

# Recap &mdash; spatio-temporal

```{r biom-space-time-plot, fig.height=8, fig.width=15, echo=FALSE}
coast <- read_sf(here("data/nl_coast.shp"))
ggplot(shrimp) +
  geom_point(aes(x = long, y = lat, size = shrimp), alpha=.5) +
  geom_sf(data = coast) +
  facet_wrap(~year, ncol = 5)
```
---

# Recap &mdash; spatio-temporal model

```{r fit-shrimp-space-time}
m_spt <- gam(shrimp ~ offset(log(area_trawled)) +
                 te(x, y, year, d = c(2,1), bs = c('tp', 'cr'), k = c(20, 5)),
             data = shrimp,
             family = tw,
             method = "REML")
```

---

# Confidence bands

.row[
.col-6[

`plot()`

```{r plot-richness-model}
plot(m_rich)
```
]
.col-6[

`gratia::draw()`

```{r draw-richness-model}
draw(m_rich)
```
]
]

What do the bands represent?

---

# Confidence intervals for smooths

Bands are a bayesian 95% credible interval on the smooth

`plot.gam()` draws the band at &plusmn; **2** std. err.

`gratia::draw()` draws them at $(1 - \alpha) / 2$ upper tail probability quantile of $\mathcal{N}(0,1)$

`gratia::draw()` draws them at ~ &plusmn;**1.96** std. err. & user can change $\alpha$ via argument `ci_level`

--

So `gratia::draw()` draws them at ~ &plusmn;**2** st.d err

---

# Across the function intervals

The *frequentist* coverage of the intervals is not pointwise &mdash; instead these credible intervals have approximately 95% coverage when *averaged* over the whole function

Some places will have more than 95% coverage, other places less

--

Assumptions yielding this result can fail, where estimated smooth is a straight line

--

Correct this with `seWithMean = TRUE` in `plot.gam()` or `overall_uncertainty = TRUE` in `gratia::draw()`

This essentially includes the uncertainty in the intercept in the uncertainty band

---

# Correcting for smoothness selection

The defaults assume that the smoothness parameter(s) $\lambda_j$ are *known* and *fixed*

--

But we estimated them

--

Can apply a correction for this extra uncertainty via argument `unconditional = TRUE` in both `plot.gam()` and `gratia::draw()`

---

# But still, what do the bands represent?

```{r plot-conf-band-plus-posterior-smooths, fig.height = 5}
sm_fit <- evaluate_smooth(m_rich, 's(year)') # tidy data on smooth
sm_post <- smooth_samples(m_rich, 's(year)', n = 20, seed = 42) # more on this later
draw(sm_fit) + geom_line(data = sm_post, aes(x = .x1, y = value, group = draw),
                         alpha = 0.3, colour = 'red')
```

---
class: inverse middle center subsection

# Your turn!

---
class: inverse middle center subsection

# Using your shiny new model

---

# Predicting with `predict()`

`plot.gam()` and `gratia::draw()` show the component functions of the model on the link scale

Prediction allows us to evaluate the model at known values of covariates on the response scale

Use the standard function `predict()`

Provide `newdata` with a data frame of values of covariates

---

# `predict()`

```{r predict-newdata}
new_year <- with(shrimp, tibble(year = seq(min(year), max(year), length.out = 100)))
pred <- predict(m_rich, newdata = new_year, se.fit = TRUE, type = 'link')
pred <- bind_cols(new_year, as_tibble(as.data.frame(pred)))
pred
```

---

# `predict()` &rarr; response scale

```{r predict-newdata-resp}
ilink <- inv_link(m_rich)                         # inverse link function
crit <- qnorm((1 - 0.89) / 2, lower.tail = FALSE) # or just `crit <- 2`
pred <- mutate(pred, richness = ilink(fit),
               lwr = ilink(fit - (crit * se.fit)), # lower...
               upr = ilink(fit + (crit * se.fit))) # upper credible interval
pred
```

---

# `predict()` &rarr; plot

Tidy objects like this are easy to plot with `ggplot()`

```{r plot-predictions-richness, fig.height = 4}
ggplot(pred, aes(x = year)) +
    geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2) +
    geom_line(aes(y = richness)) + labs(y = "Species richness", x = NULL)
```

---

# `predict()` for space and time

This idea is very general;  spatiotemporal model needs a grid of x,y coordinates for each year

```{r spt-example-predict}
sp_new <- with(shrimp, expand.grid(x = seq_min_max(x, n = 100), y = seq_min_max(y, n = 100),
                                   year = unique(year), area_trawled = median(area_trawled)))
sp_pred <- predict(m_spt, newdata = sp_new, se.fit = TRUE) # link scale is default
sp_pred <- bind_cols(as_tibble(sp_new), as_tibble(as.data.frame(sp_pred)))
sp_pred
```

---

# `predict()` &rarr; response scale

```{r spt-example-response-scale}
ilink <- inv_link(m_spt)
too_far <- exclude.too.far(sp_pred$x, sp_pred$y, shrimp$x, shrimp$y, dist = 0.1)
sp_pred <- sp_pred %>% mutate(biomass = ilink(fit),
                              biomass = case_when(too_far ~ NA_real_,
                                                  TRUE ~ biomass))
sp_pred
```

---

# `predict()` &rarr; plot

```{r spt-example-plot, fig.height = 5.5}
ggplot(sp_pred, aes(x = x, y = y, fill = biomass)) + geom_raster() +
    scale_fill_viridis_c(option = "plasma") + facet_wrap(~ year, ncol = 5) + coord_equal()
```

---

# Visualizing the trend?

We have this model

.smaller[
```{r show-m-spt}
m_spt
```
]

How would you visualize the average change in biomass over time?

---

# Welcome back old friend

One way is to  decompose the spatio-temporal function in main effects plus interaction

```{r shrimp-ti-model}
m_ti <- gam(shrimp ~ offset(log(area_trawled)) +
                ti(x, y, year, d = c(2, 1), bs = c("tp", "cr"), k = c(20, 5)) +
                s(x, y, bs = "tp", k = 20) +
                s(year, bs = "cr", k = 5),
            data = shrimp, family = tw, method = "REML")
```

and predict from the model using only the marginal effect of `s(year)`

---

# `predict()` with `exclude`

.row[
.col-6[
We can exclude the spatial & spatiotemporal terms from predictions using `exclude`

**Step 1** run `summary()` on model & note the names of the smooth you *don't* want &rarr;
]
.col-6[
.smaller[
```{r summary-spt-ti}
summary(m_ti)
```
]
]
]

---

# `predict()` with `exclude` &mdash; Step 2 *predict*

Prediction data only need dummy values for `x` and `y`

```{r pred-data-ti-model}
ti_new <- with(shrimp, expand.grid(x = mean(x), y = mean(y), year = seq_min_max(year, n = 100),
                                   area_trawled = median(area_trawled)))

{{ti_pred <- predict(m_ti, newdata = ti_new, se.fit = TRUE, exclude = c("ti(x,y,year)", "s(x,y)"))}}

ti_pred <- bind_cols(as_tibble(ti_new), as_tibble(as.data.frame(ti_pred))) %>%
    mutate(biomass = ilink(fit),
           lwr = ilink(fit - (crit * se.fit)),
           upr = ilink(fit + (crit * se.fit)))
```

`exclude` takes a character vector of terms to exclude &mdash; `predict()` sets the contributions of those terms to 0

Could also use `terms = "s(year)"` to select only the named smooths

```{r pred-data-ti-model-terms, results = "hide"}
predict(m_ti, newdata = ti_new, se.fit = TRUE, terms = "s(year)")
```

---

# `predict()` with `exclude`&mdash; Step 3 *plot it!*

```{r plot-ti-marginal-trend, fig.height = 5}
ggplot(ti_pred, aes(x = year)) + geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.3) +
    geom_line(aes(y = biomass)) + labs(y = "Biomass", x = NULL)
```

---
class: inverse middle center subsection

# Your turn!

---
class: inverse middle center subsection

# Posterior simulation

---

# Remember this?

```{r plot-conf-band-plus-posterior-smooths, fig.height = 5, echo = FALSE}
```

--

Where did the red lines come from?

---

# Posterior distributions

Each red line is a draw from the *posterior distribution* of the smooth

Remember the $\beta_j$ from Eric's opening lecture yesterday?

Together they are distributed *multivariate normal* with

* mean vector given by $\hat{\beta}_j$
* covariance matrix $\boldsymbol{\hat{V}}_{\beta}$

$$\text{MVN}(\boldsymbol{\hat{\beta}}, \boldsymbol{\hat{V}}_{\beta})$$

--

The model as a whole has a posterior distribution too

--

We can simulate data from the model by taking draws from the posterior distribution

---

# Posterior simulation for a smooth

Sounds fancy but it's only just slightly more complicated than using `rnorm()`

To do this we need a few things:

1. The vector of model parameters for the smooth, $\boldsymbol{\hat{\beta}}$
2. The covariance matrix of those parameters, $\boldsymbol{\hat{V}}_{\beta}$
3. A matrix $\boldsymbol{X}_p$ that maps parameters to the linear predictor for the smooth

$$\boldsymbol{\hat{\eta}}_p = \boldsymbol{X}_p \boldsymbol{\hat{\beta}}$$

--

Let's do this for `m_rich`

---

# Posterior sim for a smooth &mdash; step 1

The vector of model parameters for the smooth, $\boldsymbol{\hat{\beta}}$

```{r richness-coefs}
sm_year <- get_smooth(m_rich, "s(year)") # extract the smooth object from model
idx <- gratia:::smooth_coefs(sm_year)    # indices of the coefs for this smooth
idx

beta <- coef(m_rich)                     # vector of model parameters
```

---

# Posterior sim for a smooth &mdash; step 2

The covariance matrix of the model parameters, $\boldsymbol{\hat{V}}_{\beta}$

```{r richness-vcov}
Vb <- vcov(m_rich) # default is the bayesian covariance matrix
```
---

# Posterior sim for a smooth &mdash; step 3

A matrix $\boldsymbol{X}_p$ that maps parameters to the linear predictor for the smooth

We get $\boldsymbol{X}_p$ using the `predict()` method with `type = "lpmatrix"`

```{r richness-xp-matrix}
new_year <- with(shrimp, tibble(year = seq_min_max(year, n = 100)))
Xp <- predict(m_rich, newdata = new_year, type = 'lpmatrix')
dim(Xp)
```

---

# Posterior sim for a smooth &mdash; step 4

Take only the columns of $\boldsymbol{X}_p$ that are involved in the smooth of `year`

```{r richness-reduce-xp}
Xp <- Xp[, idx, drop = FALSE]
dim(Xp)
```

---

# Posterior sim for a smooth &mdash; step 5

Simulate parameters from the posterior distribution of the smooth of `year`

```{r richness-simulate-params}
set.seed(42)
beta_sim <- rmvn(n = 20, beta[idx], Vb[idx, idx, drop = FALSE])
dim(beta_sim)
```

Simulating many sets (20) of new model parameters from the estimated parameters and their uncertainty (covariance)

Result is a matrix where each row is a set of new model parameters, each consistent with the fitted smooth

---

# Posterior sim for a smooth &mdash; step 6

.row[
.col-6[
Form $\boldsymbol{\hat{\eta}}_p$, the posterior draws for the smooth

```{r richness-posterior-draws, fig.height = 5, fig.show = 'hide'}
sm_draws <- Xp %*% t(beta_sim)
dim(sm_draws)
matplot(sm_draws, type = 'l')
```

A bit of rearranging is needed to plot with `ggplot()`
]

.col-6[
```{r richness-posterior-draws, fig.height = 5, fig.width = 5, echo = FALSE, results = 'hide'}
```
]

]

--

Or use `smooth_samples()`

---

# Posterior sim for a smooth &mdash; steps 1&ndash;6

```{r plot-posterior-smooths, fig.height = 5}
sm_post <- smooth_samples(m_rich, 's(year)', n = 20, seed = 42)
draw(sm_post)
```

---

# Posterior simulation from the model

Simulating from the posterior distribution of the model requires 1 modification of the recipe for a smooth and one extra step

We want to simulate new values for all the parameters in the model, not just the ones involved in a particular smooth

--

Additionally, we could simulate *new response data* from the model and the simulated parameters (**not shown** below)

---

# Posterior simulation from the model

```{r posterior-sim-model}
beta <- coef(m_rich)   # vector of model parameters
Vb <- vcov(m_rich)     # default is the bayesian covariance matrix
Xp <- predict(m_rich, type = 'lpmatrix')
set.seed(42)
beta_sim <- rmvn(n = 1000, beta, Vb) # simulate parameters
eta_p <- Xp %*% t(beta_sim)        # form linear predictor values
mu_p <- inv_link(m_rich)(eta_p)    # apply inverse link function

mean(mu_p[1, ]) # mean of posterior for the first observation in the data
quantile(mu_p[1, ], probs = c(0.025, 0.975))
```

---

# Posterior simulation from the model

```{r posterior-sim-model-hist, fig.height = 5}
ggplot(tibble(richness = mu_p[587, ]), aes(x = richness)) +
    geom_histogram() + labs(title = "Posterior richness for obs #587")
```

---

# Posterior simulation from the model

Or easier using `fitted_samples()`

```{r richness-fitted-samples, fig.height = 4.5}
rich_post <- fitted_samples(m_rich, n = 1000, newdata = shrimp, seed = 42)
ggplot(filter(rich_post, row == 587), aes(x = fitted)) +
    geom_histogram() + labs(title = "Posterior richness for obs #587", x = "Richness")
```

---

# Why is this of interest?

Say you wanted to get an estimate for the total biomass of shrimp over the entire region of the trawl survey for 2007

You could predict for the spatial grid for `year == 2007` using code shown previously and sum the predicted biomass values over all the grid cells

--

**Easy**

--

But what if you also wanted the uncertainty in that estimate?

--

**Hard**

--

**Math**  😱😱😱 "something, something, delta method, something" 😱😱😱 

---

# Posterior simulation makes this easy

1. Take a draw from the posterior distribution of the model
2. Use the posterior draw to predict biomass for each grid cell
3. Sum the predicted biomass values over all grid cells
4. Store the total biomass value
5. Repeat 1&ndash;4 a lot of times to get posterior distribution of total biomass
6. Summarize the total biomass posterior
    * Estimated total biomass is the mean of the total biomass posterior
	* Uncertainty is some lower/upper tail probability quantiles of the posterior

---

# Let's do it

```{r total-biomass-posterior-1}
sp_new <- with(shrimp, expand.grid(x = seq_min_max(x, n = 100), y = seq_min_max(y, n = 100),
                                   year = 2007, area_trawled = median(area_trawled)))
Xp <- predict(m_spt, newdata = sp_new, type = "lpmatrix")

## work out now which points are too far now
too_far <- exclude.too.far(sp_new$x, sp_new$y, shrimp$x, shrimp$y, dist = 0.1)

beta <- coef(m_spt)   # vector of model parameters
Vb <- vcov(m_spt)     # default is the bayesian covariance matrix
set.seed(42)
beta_sim <- rmvn(n = 1000, beta, Vb) # simulate parameters
eta_p <- Xp %*% t(beta_sim)        # form linear predictor values
mu_p <- inv_link(m_spt)(eta_p)     # apply inverse link function
```

Columns of `mu_p` contain the expected or mean biomass for each grid cell per area trawled

Sum the columns of `mu_p` and summarize

---

# Summarize the expected biomass

```{r total-biomass-posterior-2, dependson = -1}
mu_copy <- mu_p              # copy mu_p
mu_copy[too_far, ] <- NA     # set cells too far from data to be NA
total_biomass <- colSums(mu_copy, na.rm = TRUE)  # total biomass over the region

mean(total_biomass)
quantile(total_biomass, probs = c(0.025, 0.975))
```

---

# Summarize the expected biomass

```{r total-biomass-histogram, echo = FALSE}
ggplot(tibble(biomass = total_biomass), aes(x = biomass)) +
    geom_histogram()
```

---

# With `fitted_samples()`

.row[

.col-7[
```{r biomass-fitted-samples-example}
bio_post <- fitted_samples(m_spt, n = 1000,
                           newdata = sp_new[!too_far, ],
                           seed = 42) %>%
    group_by(draw) %>%
    summarise(total = sum(fitted),
              .groups = "drop_last")

with(bio_post, mean(total))
with(bio_post, quantile(total, probs = c(0.025, 0.975)))
```
]

.col-5[
```{r biomass-fitted-samples-plot, fig.width = 5, fig.height = 5}
ggplot(bio_post, aes(x = total)) +
    geom_histogram() +
    labs(x = "Total biomass")
```

]

]

---
class: inverse middle center subsection

# Your turn!

